The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-08-06 paper/iphas-dr2-rev1.tex
Revision submitted 2014-08-06!

{\bf When the previous release -- the IDR -- was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release, 
 which supercedes the IDR by including a global re-calibration
and by taking the coverage up to 92 per cent of the survey area.}
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

An external validation of the astrometry 
has been carried out by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the offsets against UCAC4
computed for 1.3 million stars
in the magnitude range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean position offset to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.
%We note that both UCAC4 and 2MASS -- and hence also IPHAS --
%not 100% true due to bulk pm of calibrators - Mike
%are calibrated in the International Celestial Reference System (ICRS).
%We note that the mean residual is  
%is $-0.006 \pm 0.086$ in R.A.
%and $0.012\pm0.073$ in Dec.

\begin{figure}
    \includegraphics[width=\linewidth]{figures/astrometry/astrometry.pdf} 
    \caption{Distribution of the position offsets
    		of stars
    		which appear both in IPHAS DR2 and UCAC4
    		within a cross-matching distance of 1\arcsec.
            The residuals were computed for 
            the 1.3 million stars in the IPHAS catalogue
            which are not blended, not saturated,
            and fall in the magnitude range $13 < r < 15$.
            The mean and standard deviation of this distribution 
            equals $0\farcs 09\pm0\farcs 07$.}
    \label{fig:astrometry}
\end{figure}

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
{ \bf
This is demonstrated in Fig.~\ref{fig:ccd_before},
where we show the combined colour-colour diagram
for nearly 3,000 fields across an area of 400~deg$^2$.
The main locus of stars is poorly defined
in the diagram due to the presence of
incorrectly calibrated fields,
which need to be corrected
during a global re-calibration step.
The application of such a procedure,
to be explained below,
has revealed that the error in our initial nightly calibration
exceeded 0.1~mag in 12 per cent of the fields,
and even exceeded 0.5~mag in 0.7 per cent.
For comparison, Fig.~\ref{fig:ccd_after}
demonstrates the improvement 
in the colour-colour diagram
after re-calibrating. }

@@

%\newpage

@@

First, Table~\ref{tab:a10} specifies
a set of minimum quality criteria
which should benefit most applications
which desire accurate colours
as well as completeness down to $\sim$19th magnitude.
The listed criteria are designed
to ensure that each band offers photometry
at a high S/N which is consistent
across different aperture sizes.
This latter criterion is a proxy
for identifying objects which are affected
by poor background subtraction
or failed deblending.
A total of 86 million sources out of 219 million 
(39 per cent) satisfy all the criteria listed in Table~\ref{tab:a10}
and are hereafter referred to as `a10'
(short for ``all-band $10\sigma$'').
For convenience, the catalogue contains a boolean column
named \emph{a10} that flags these objects.

@@

Fig.~\ref{fig:magdist} compares the $r$-band magnitude
distributions for objects tagged \emph{a10} and \emph{a10point}
against the unfiltered catalogue.
In the magnitude range $13 < r < 19$,
we find that 81 per cent of the sources 
are flagged \emph{a10}
and 54 per cent are flagged \emph{a10point}.
We will explain below that the
\emph{a10point} category is least complete
at low Galactic longitudes,
where source blending can affect
up to a quarter of the objects.

@@

Having obtained a set of quality criteria,
we now assess how these improve the data quality.
In this section we start by discussing
the \emph{reliability} of the survey,
which we define as the fraction of sources
which are genuine astrophysical objects.
To estimate the reliability, we cross-matched IPHAS
against SDSS DR7\footnote{SDSS DR7 was chosen for this comparison
due to an apparent change in 
the catalogue generation for DR8 and DR9,
which resulted in fewer genuine stars being
detected near bright ($r\,<\,12$) sources,
which in turn gave an unduly pessimistic view
on our reliability.},
which is a catalogue of similar resolution and depth.
By computing the fraction of IPHAS sources which were
also independently detected by SDSS,
we can obtain a good lower limit on the reliability.

The comparison was carried out across three cone-shaped regions,
each covering 1 deg$^2$,
using a strict cross-matching distance of 0.5~arcsec.
We considered four subsets of the IPHAS catalogue:
\begin{enumerate}
\item the entire (unfiltered) catalogue;
\item objects detected in all three bands;
\item \emph{a10} sources;
\item \emph{a10point} sources.
\end{enumerate}

@@

\begin{table}
    \caption{Fractions of sources in our catalogue
             which are also found in the SDSS DR7 catalogue
             within a cross-matching radius of 0.5~arcsec.
             These statistics can be interpreted
             as lower limits
             for the reliability of our sources,
             i.e. the probability that an entry in our
             catalogue is a genuine astrophysical source
             rather than an instrumental artefact.
             The fractions were computed
             in three 1 deg$^2$ regions
             where both surveys
             show contiguous overlap.
             The reliability is shown
             for IPHAS sources 
             without selection criteria applied (``all''),
             using only objects detected in all three bands ($nBands=3$),
             using objects detected at 10$\sigma$
             in all three bands (\emph{a10}),
             and finally using the detections
             classified as reliable 10$\sigma$ point sources (\emph{a10point}).
             }
    \label{tbl:reliability}
    \begin{center}
        \begin{tabular}{lcccccc}
        \toprule
        Region ($\ell$, $b$) & all & $nBands=3$ & \emph{a10} & \emph{a10point} \\
        \midrule
        (149.39, 4.06) & 93.0\% & 98.8\% & 99.7\% & 99.8\% \\
        (186.59, -2.50) & 90.0\% & 98.8\% & 99.4\% & 99.6\% \\
        (202.70, -1.75) & 89.4\% & 98.5\% &99.4\% & 99.7\% \\ \midrule
        mean & 90.8\% & 98.7\% & 99.5\% & 99.7\% \\
        \bottomrule
        \end{tabular}
    \end{center}
\end{table}

@@

{\bf Towards Sh2-82,
the IPHAS and Spitzer/GLIMPSE catalogues
have 4798 entries in common,
out of 10\,739 and 11\,321 entries in total, respectively.
Fig.~\ref{fig:spitzer-ccd} presents the Spitzer colour-colour diagram
of the region, containing the 1356 objects that were detected 
in all four GLIMPSE bands.
We have indicated the 845 objects (62 per cent)
that have a counterpart in IPHAS (grey circles)
and the 511 objects that do not (red crosses).

@@

We find that objects located in the upper half of the Spitzer diagram
are less likely to have an IPHAS counterpart,
which is where highly-reddened objects are expected to sit.
This is not surprising,
because Sh2-82 is located 
towards a high-extinction sightline 
near the inner plane,
where an optical survey will naturally be extinction-limited.
Nevertheless, the extinction map due to \citet{Sale2014}
has demonstrated that IPHAS can probe stars as far away 
as 4-5\,kpc towards this region.
The situation will be even more favourable
in the outer-plane, where dust optical depths are low.

@@

%where IPHAS is not limited by extinction
%This is an example of a high-extinction sightline where
%optical observations will naturally be extinction-limited.
%IPHAS will naturally count far fewer sources than (N)IR surveys.
%In the outer-plane the situation is, in fact, reversed,
%except where there is UKIDSS GPS coverage (i.e. dec > 60).
%For example, in the anti-centre region where the dust optical depths are low, IPHAS contains more...

%Amongst the 1810 high-quality sources which we
%plotted in Fig.~\ref{fig:emitters},
%no less than 1561 objects (87\%) have a counterpart
%in the Spitzer GLIMPSE archive
%within a matching radius of 1 arcsec.
%In the other direction, out of the 6753 objects
%which are in the Spitzer/GLIMPSE catalogue for this region,
%2984 (44\%) have a counterpart in IPHAS.

% reliable & r < 20 & (l > 53.55-10/60.) & (l < 53.55+10/60.)                  & (b > 0.01-7.5/60.) & (b < 0.01+7.5/60.)
%TODO: This is an example of a high-extinction sightline where
% optical observations will naturally be extinction-limited.
% IPHAS will naturally count far fewer sources than (N)IR surveys.
% In the outer-plane the situation is, in fact, reversed,
% except where there is UKIDSS GPS coverage (i.e. dec > 60).
% For example, in the anti-centre region where the dust optical depths are low, IPHAS contains more...

@@

%\vspace{0.5cm}

>>>>

{\bf When the previous release -- the IDR -- was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release, 
which supersedes the IDR by including a global re-calibration
and by taking the coverage up to 92 per cent of the survey area.}
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

An external validation of the astrometry 
has been carried out by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the offsets against UCAC4
computed for 1.3 million stars
in the magnitude range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean position offset to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.

\begin{figure}
    \includegraphics[width=\linewidth]{figures/astrometry/astrometry.pdf} 
    \caption{Distribution of the astrometric residuals of stars
    		which appear both in IPHAS DR2 and UCAC4
    		within a cross-matching distance of 1\arcsec.
            The residuals were computed for 
            the 1.3 million stars in the IPHAS catalogue
            which are not blended, not saturated,
            and fall in the magnitude range $13 < r < 15$.
            The mean and standard deviation of this distribution 
            equals $0\farcs 09\pm0\farcs 07$.}
    \label{fig:astrometry}
\end{figure}

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
{ \bf
This is demonstrated in Fig.~\ref{fig:ccd_before},
where we show the combined colour-colour diagram
for nearly 3,000 fields across an area of 400~deg$^2$.
The main locus of stars is poorly defined
in the diagram due to the presence of
incorrectly calibrated fields,
which need to be corrected
during a global re-calibration step.
The application of such a procedure,
to be explained below,
has revealed that the error in our initial nightly calibration
exceeded 0.1~mag in 12 per cent of the fields,
and even exceeded 0.5~mag in 0.7 per cent.
Fig.~\ref{fig:ccd_after}
demonstrates the improvement 
in the colour-colour diagram
after re-calibrating. }

@@

@@

First, Table~\ref{tab:a10} specifies
a set of minimum quality criteria
which should benefit most applications
which desire accurate colours
as well as completeness down to $\sim$19th magnitude.
The listed criteria are designed
to ensure that each band offers photometry
at S/N\ $>10$ that is self-consistent
across different aperture sizes.
%This latter criterion is a proxy
%for identifying objects which are affected
%by poor background subtraction
%or failed deblending.
A total of 86 million sources out of 219 million 
(39 per cent) satisfy all the criteria listed in Table~\ref{tab:a10}
and are hereafter referred to as `a10'
(short for ``all-band $10\sigma$'').
For convenience, the catalogue contains a boolean column
named \emph{a10} that flags these objects.

@@

Fig.~\ref{fig:magdist} compares the $r$-band magnitude
distributions for objects tagged \emph{a10} and \emph{a10point}
against the unfiltered catalogue.
we find that 81 per cent of sources 
in the magnitude range $13 < r < 19$
are flagged \emph{a10},
and 54 per cent are flagged \emph{a10point}.
We will show below that the
\emph{a10point} category is least complete
at low Galactic longitudes,
where source blending can affect
up to a quarter of the objects.

@@

Having obtained a set of quality criteria,
we now assess how these improve the data quality.
In this section we start by discussing
the \emph{reliability} of the survey,
which we define as the fraction of sources
which are genuine astrophysical objects.
To estimate the reliability, we cross-matched IPHAS
against SDSS DR7\footnote{SDSS DR7 was chosen
due to an apparent change in 
the catalogue preparation for DR8 and DR9,
which resulted in fewer genuine stars being
detected near bright ($r\,<\,12$) sources,
which in turn gave an unduly pessimistic view
on our reliability.},
which is a catalogue of similar resolution and depth.
By computing the fraction of IPHAS sources which were
also independently detected by SDSS,
we can obtain a good lower limit on the reliability.

The comparison was carried out across three cone-shaped regions,
each covering 1 deg$^2$,
using a strict cross-matching distance of 0\farcs 5.
We considered four subsets of the IPHAS catalogue:
\begin{enumerate}
\item the entire (unfiltered) catalogue;
\item objects detected in all three bands;
\item \emph{a10} sources;
\item \emph{a10point} sources.
\end{enumerate}

\begin{table}
    \caption{Fractions of sources in our catalogue
             which are also found in SDSS DR7
             within a cross-matching radius of 0\farcs 5.
             These percentages can be interpreted
             as lower limits
             for the reliability of our sources,
             i.e. the probability that an entry in our
             catalogue is a genuine astrophysical source
             rather than an instrumental artefact.
             The fractions were computed
             in three 1 deg$^2$ regions
             where both surveys
             show contiguous overlap.
             The reliability is shown
             without quality criteria applied (``all''),
             using only objects detected in all three bands ($nBands=3$),
             using objects detected at 10$\sigma$
             in all three bands (\emph{a10}),
             and finally using objects
             classified as reliable 10$\sigma$ point sources (\emph{a10point}).
             }
    \label{tbl:reliability}
    \begin{center}
        \begin{tabular}{lcccccc}
        \toprule
        Region ($\ell$, $b$) & all & $nBands=3$ & \emph{a10} & \emph{a10point} \\
        \midrule
        (149.39, 4.06) & 93.0\% & 98.8\% & 99.7\% & 99.8\% \\
        (186.59, -2.50) & 90.0\% & 98.8\% & 99.4\% & 99.6\% \\
        (202.70, -1.75) & 89.4\% & 98.5\% &99.4\% & 99.7\% \\ \midrule
        mean & 90.8\% & 98.7\% & 99.5\% & 99.7\% \\
        \bottomrule
        \end{tabular}
    \end{center}
\end{table}

@@

@@

{\bf Towards Sh2-82,
the IPHAS and Spitzer/GLIMPSE catalogues
have 4,798 entries in common,
out of 10,739 and 11,321 entries in total, respectively.
Fig.~\ref{fig:spitzer-ccd} presents the Spitzer colour-colour diagram
of the region, showing the 1,356 objects that were detected 
in all four GLIMPSE bands.
We have indicated the 845 objects (62 per cent)
that have a counterpart in IPHAS (grey circles)
and the 511 objects that do not (red crosses).

@@

We find that objects located in the upper half of the Spitzer diagram
are less likely to have an IPHAS counterpart,
which is where highly-reddened objects are expected to sit.
Moreover, amongst the objects with a counterpart,
we note that more have been detected in $i$ (843 objects)
than in $r$ (634) or \ha\ (600).
This is not surprising,
because Sh2-82 is located 
towards a high-extinction sightline 
near the inner plane,
where an optical survey will naturally be extinction-limited.
Nevertheless, the extinction map due to \citet{Sale2014}
has demonstrated that IPHAS can probe stars as far away 
as 4-5\,kpc towards this region.
The situation will be even more favourable
in the outer-plane, where dust optical depths are low.

@@

@@
